---
title: "Instacart Reorder Analysis and Prediction"
author: Gwyndolyn Bleikamp
date: 08/01/2017
output: html_notebook
---

# Yapyapyap

## Objective
Instacart would like to be able to predict for the test set (orders.csv)

1. Products that a user will buy again,
2. Try for the first time, or 
3. Add to cart next during a session

## Data
Data is provided open-source online 
(https://www.instacart.com/datasets/grocery-shopping-2017)

There is also a kaggle competition to enter for a chance to win a prize
(https://www.kaggle.com/c/instacart-market-basket-analysis/data)

Data provided are for each prior/training/testing sets:  
* aisle 
* department (Frozen, Bakery, etc...)
* product name
* Indicator for a reordered product
* Day of week/Time of order

Submission File Format:  
For each order ID in the test set, provide the following  
Order ID,Space delimited list of products or the word 'None'   
17,1 2
34,None
...

## Load Necessary Libraries
```{r, message = FALSE}
# Necessary Evils
library(dplyr)
library(ggplot2)
library(readr)
```


# Exploratory Data Analysis

## Dimension Level

First read in the data
```{r}

# Read in dimension level data
aisles <- read.csv("C:/Users/owner/Documents/Kaggle/Instacart/Instacart_Data/instacart_2017_05_01/aisles.csv",
                   stringsAsFactors = FALSE)
depts  <- read.csv("C:/Users/owner/Documents/Kaggle/Instacart/Instacart_Data/instacart_2017_05_01/departments.csv",
                   stringsAsFactors = FALSE)
products  <- read.csv("C:/Users/owner/Documents/Kaggle/Instacart/Instacart_Data/instacart_2017_05_01/products.csv",
                      stringsAsFactors = FALSE)

# Merge in aisle and department descriptions with the product by id
product_dim <- left_join(products, aisles, by = "aisle_id") %>%
  left_join(., depts, by = "department_id")

```

Explore a bit with the dimensional data before reading in event level purchases
```{r}
str(product_dim)
```

There are 49,688 distinct products. Now we have aisle and dept mapped in so 
explore each one at a time to get our feet wet.

### Departments

First look at which departments have the most products
```{r}
# Get count by department and produce bar plot sorted
product_dim %>%
  group_by(department) %>%
  summarize(Frequency = length(department)) %>%
  mutate(department = factor(department, 
                             levels = unique(department[order(Frequency)]))) %>%
  
  ggplot(data = ., aes(x = department, y = Frequency)) + 
    geom_bar(stat = "identity") + 
    coord_flip() + 
    labs(title = "Count of Products within each Department")
```

Personal care has over 6000 products while < 2,000 different types of produce exist. I could guess there are multiple brands and types of hair spray but there is only one 'lettuce'. Personal care is a bit more broad of a category as well. How many aisles could consititue personal care?

```{r}
as.matrix(table(product_dim[product_dim$department == "personal care",]$aisle))
```
Over 1,000 types of vitamins! Plus over 800 products in 'hair care' aisles 
across all represented retailers (unknown).

Get a feel for the granularity of the product - Are there more than one hairspray? 
```{r}
filter(product_dim, aisle == "hair care" & grepl("spray", product_name))
```

What about number of times bought. Is the generic "Hairspray" (34192) more
common since there are only 29 other distinct "hair care" aisle prodcuts 
with the word 'spray' in the name?


How granular do tomatoes get? Cherry/organic/yellow/red/garden/roma/etc...
```{r}
filter(product_dim, department == "produce" & grepl("tomato", tolower(product_name)))
```
Answer - Very!
Is there a skew towards generic product name (Tomato)?

## Fact Level: Training Data
Bring in order training/prior and understand how they work together

```{r}
# Fact level data:
# Training Set of orders
orders_train <- read_csv("C:/Users/owner/Documents/Kaggle/Instacart/Instacart_Data/instacart_2017_05_01/order_products__train.csv")
```

What does this data look like?
```{r}
str(orders_train)
```
Join on product id to the product_dim table
```{r}
fact_train <- left_join(orders_train,
                        product_dim,
                        by = "product_id")

str(fact_train)
```

How many orders are in the training data?
```{r}
length(unique(fact_train$order_id))
```

What items are most popular?
```{r}

# How many of the ~50K products are bought in the training set
length(unique(fact_train$product_name))

# Get a feel for the dist of Frequency
freq_df <- fact_train %>%
  group_by(product_name) %>%
  summarize(Frequency = length(product_name)) %>%
  mutate(product_name = factor(product_name, 
                               levels = unique(product_name[order(Frequency)]))) 
quantile(freq_df$Frequency, c(0, .25, .5, .75, .95, 1))
hist(freq_df$Frequency)

# Get count of products, ordered by frequency
# So many to graph them all so just choose the top 20 products
fact_train %>%
  group_by(product_name) %>%
  summarize(Frequency = length(product_name)) %>%
  mutate(product_name = factor(product_name, 
                               levels = unique(product_name[order(Frequency)]))) %>% 
  top_n(20) %>%
  
  ggplot(data = ., aes(x = product_name, y = Frequency)) + 
    geom_bar(stat = "identity") + 
    coord_flip() + 
    labs(title = "Top 20 Products bought in the Training set")
```

Next Steps
Parse out organic vs not indicator
Link similar products (Avocado vs Hass Avocado) - Do they mind which is which?
Perhaps we can upsell to a more expensize avocado.
Look at the top departments/aisles - The top 20 products all seem to be produce.
It seems that perhaps our product list is not every product, but rather any product that was bought - Could that introduce bias? We will never recommend products that have never been bought before?